The brightness and spatial distributions of terrestrial radio sources

### 2013-01-07 articles/Article-RFI-distribution/article-rfi-distribution.tex
First part of Garrelts comments

\begin{abstract}
Feeble undetected sources of radio-frequency interference (RFI) might become visible after integration of long radio observations if they are coherently present over time. Thereby, they might obstruct the detection of the astronomical signal of interest. This issue is especially important for Epoch of Reionisation (EoR) projects that try to detect the faint redshifted HI signals of the earliest objects in the Universe. We explore the low-frequency RFI situation by studying the $\log N$--$\log S$ histograms of data observed with LOFAR. The histograms shows an RFI distribution that follows a power-law distribution with an estimated exponent around -1.6 to -1.5. With several assumptions, this can be explained with a uniform distribution of terrestrial radio sources whose radiation follows existing propagation models. Extrapolation of the power law would imply that the current LOFAR EoR observations should be severely RFI limited if these sources are coherently present. Since these observations have achieved the thermal (sky) noise limit, a possible implication is that most of the RFI sources do not add coherently.
\end{abstract}

@@

\section{Introduction}
Radio astronomy concerns itself with the observation of radiation from celestial sources at radio wavelengths. However, astronomical radio observations can be affected by radio-frequency interference (RFI), which makes it difficult to calibrate the instrument and achieve high sensitivities \citep{impact-of-warc79,interference-and-radioastronomy-1991,interference-model-lemmon,rfi-mitigation-overview-fridman-baan}. Many techniques have been designed to mitigate this effect, such as detection and flagging of data \citep{chi-square-time-blanking-weber, multichannel-rfi-mitigation, exoplanet-detection-with-rfi, wsrt-rfims, pulse-blanking, effelsberg-rfi-mitigation, LOFAR-RFI-pipeline}, adaptive cancellation techniques \citep{adaptive-cancellation,post-correlation-reference-signal} and spatial filtering \citep{multichannel-rfi-mitigation, ellingson-spatial-nulling-2002, hampson-spatial-nulling-2002, boonstra-dissertation, spatial-filtering-parkes-multibeam-for-pulses, post-correlation-filtering}. In most cases, these methods excise enough of the interference to calibrate and image the data.

Typical radio observations record a few hours of data, and the results are integrated. In these cases, it often suffices to only excise interference that is apparent and thus above the noise to reach the thermal noise limit of the instrument. A new challenge arises, however, when one desires much deeper observations, and hundreds of hours of observations need to be integrated. In such a case, feeble interference caused by stationary sources might not manifest itself above the noise in individual observations, but might be coherently present in the data. Subsequently, when averaging these data, the interference might become apparent and occlude the signal of interest. This is very relevant for the 21-cm Epoch of Reionisation (EoR) experiments, because they involve long integration times. Several such experiments are underway, to either measure the angular power spectrum \citep{gmrt-eor-2011-paciga,de-bruyn-eor-ursi-2011,eor-paper-2011-jacobs,eor-mwa-2012-williams} or the global signal \citep{edges}. For these experiments, it is important to know the possible effect of low-level interferences on the data, as these might overshadow or alter the signal of interest. In this article, we will connect new insights about RFI to the angular Epoch of Reionisation experiment that is using the Low-Frequency Array \citep{de-bruyn-eor-ursi-2011}.

This work explores the information that is present in interference distributions, in order to analyse possible low-level interference that is not detectable by standard detection methods. Our approach is similar to the radio source count analysis that is used in cosmology \citep{vonhoerner-radio-source-counts}, also named $\log N$ -- $\log S$ analysis, where $N$ and $S$ refer to the celestial source count and brightness respectively. The slope in such a plot contains information about source populations, their luminosity functions and the geometry of the Universe. We analyse such a double-logarithmic plot for the case of terrestrial sources, with the ultimate goal of predicting their full spatial and brightness distributions. This results in a better insight in the effects of low-level interference and allows one to simulate the effects of interference more accurately. A distribution can be expressed either as a differential or cumulative function. This work uses differential distributions, because they are better suited for visualizing the parameters we are interested in, although the two basically convey the same information.

@@

Next, we assume that all RFI sources have equal constant strength $I$ and follow a uniform spatial distribution in a two-dimensional plane. This is obviously a simplification, as the RFI sources are actually distributed on the surface of a sphere with different heights. Beyond some distance, the Earth will partly block the path between transmitter and receiver. Therefore, the assumption of uniformly distributed sources on a two-dimensional plane is only approximately valid. Using these assumptions, we can express the expected cumulative frequency density of sources $f_I(r)$ at distance $r$ as
\begin{equation} \label{eq:cumulative-distance-function}
F_{\textrm{distance} \le r}(r)=c\pi r^2,
\end{equation}
for some constant $c$ that represents the number of sources per unit area. In other words, we will observe $F$ sources that are at most a distance $r$ away.

@@

Converting from a subtrahend in decibels to a flux density factor $L_S$, and collecting the terms of Eq.~\eqref{eq:hata-model} that are not depending on $r$ in a single variable $\zeta$, results in
\begin{equation} \label{eq:propagation-loss-factor}
 L_S = \zeta r^{4.49 - 0.655 \log_{10} h_b},
\end{equation}
with
\begin{equation}
 \zeta = \frac{f_c^{2.616}}{h_b^{1.382}} - 10^{6.955-\frac{1}{10} a(h_m)}.
\end{equation}
Note that according to Hata's model, the exponent of the power law in Eq.~\eqref{eq:propagation-loss-factor} depends only on the height of the transmitting antenna, i.e., it is independent of frequency, receiver height and urban density. Now, if in Eqs.~\eqref{eq:cumulative-rfi-distribution} and \eqref{eq:two-dimensional-distribution} one replaces the definition of $S(r)$ from Eq.~\eqref{eq:amplitude-fall-off-in-free-space} with one that includes the propagation effects,
\begin{equation}
 S(r) = \frac{\zeta Ig}{r^\eta},
\end{equation}
with $\eta = 0.655 \log_{10} h_b - 4.49$, one finds the frequency density function $f_p$ that considers propagation effects,
% S(r):=I*g/(r^alpha);
% assume(g>0,I>0,S>0);
% diff(c*%pi*r^2,S),solve(S(r)=S,r);
\begin{align}
f_p(S)
= & \frac{d}{dS} c \pi \left( \frac{\zeta Ig }{S} \right)^{2/\eta}
= \frac{c 2\pi}{\eta S} \left( \frac{\zeta Ig}{S} \right)^{2/\eta}.
\end{align}
Consequently, due to non-free space propagation effects, the observed log-log histogram is predicted to have a $\frac{2}{\eta}-1$ slope. By substituting $\eta$, one finds
\begin{equation}
\textrm{slope}(h_b) = \frac{1}{0.3275 \log_{10} h_b - 2.245} - 1.
\end{equation}
\begin{figure}
\begin{center}\hspace{-1cm}\includegraphics[width=6cm]{img/transmitter-height/plot-transmitter-height-trimmed}
\caption{Effect of transmitter height on the slope of a log-log histogram.}
\label{fig:plot-transmitter-height}
\end{center}
\end{figure}
This is valid for transmitters that have a height of 30--200 m, the range over which Hata's model was defined. This yields estimated distribution slopes of $-1.57$ and $-1.67$ for $30$ m and $200$ m high transmitters respectively. In Figure~\ref{fig:plot-transmitter-height}, the slope value is plotted as a function of the transmitter height, including extrapolated values for transmitter heights down to 1 m.

@@

To calculate the corresponding amplitude distribution that is formed by the combination of the real and imaginary component distributions, it is easier to perform numerical simulations, as analytic derivation is not trivial. The histogram can be numerically estimated by drawing complex real and imaginary samples from the two distributions and calculating and counting the amplitudes. A sample can be drawn from the RFI distribution by normalizing and inverting the cumulative frequency function in Eq.~\eqref{eq:cumulative-rfi-distribution} and evaluating it for a uniformly distributed variable. Note however that $F_{\textrm{amplitude}\le S}(S)$ is not limited; when decreasing the amplitude $S$ towards zero, the number of sources will go to infinity. Therefore, to draw a sample by using the inversion method, this would require a uniformly distributed variable $\sim U(0,\infty)$. By assuming that there are no samples beyond some limiting distance $r_L$, then $F_{\textrm{amplitude}\le S}(S) = N-c \pi r_L^2$ for $S < Ig/r_L^2$. Consequently, by sampling $x_u$ from a uniform distribution with $0 \le x_u \le N-c \pi r_L^2$ and applying the inverse of Eq.~\eqref{eq:cumulative-rfi-distribution}, we will sample amplitudes $S$ with $S \ge Ig/r_L^2$, and $S\sim$ the power-law distribution.

@@

\subsection{Determining RFI distribution limits} \label{sec:rfi-distribution-constraints}
In this section we will show methods to find $S_U$ and $S_L$, the upper and lower flux limits of the power-law distribution at which the power law breaks down. Once the exponent of the power-law part of the distribution is estimated with the previously discussed techniques, the distribution can be extrapolated to find a lower flux limit. Assume that we have found a power law with exponent $\alpha$ and factor $\beta$ over an amplitude region $[S_1 ; S_2]$, resulting in the following frequency density function $g$:
\begin{equation} \label{eq:amplitude-power-law}
g(S) = \beta S^\alpha.
\end{equation}
\begin{figure}
\begin{center}
\includegraphics[width=7.5cm]{img/explanation-lower-constraint/FindingCounts}
\includegraphics[width=7.5cm]{img/explanation-lower-constraint/FindingConstraints}
\caption{Cartoon of how a constraint on the lower fall-off point of the power-law distribution can be determined. Note that the labelled areas are areas as occupied in a linear plot, i.e., the integration of the density function. Areas in a log-log plot are not linearly related to the integral. There are two ways to estimate the lower constraint $S_L$: (i) the areas $N_a$ and $N_\textrm{total}-N_\textrm{part}$ are equal if $Ig/r^2$ is constant during the observation, and (ii) if one assumes $Ig/r^2\sim$ uniform, then $N_a+N_b = N_\textrm{total}-N_\textrm{part}$.}
\label{fig:explanation-constraints}
\end{center}
\end{figure}

Assume also that the histogram contains $N_\textrm{part}$ (RFI) samples with amplitude $> S_1$, as sketched in Fig.~\ref{fig:explanation-constraints}. The hypothetical upper limit $S_U$ of the distribution can now be found, i.e., the highest amplitude that would be observed when the distribution follows the power law up to that point, by solving
\begin{equation} \label{eq:upper-limit-constrained}
 \int\limits_{S_1}^{S_U} g(S) dS = N_\textrm{part}.
\end{equation}
In practice, the observed histogram will break down beyond some amplitude because of several reasons. First and most importantly, the samples themselves are digitized with an analogue-to-digital converter (ADC) with limited range. Second, we observe for a limited time and the frequency count is discrete. Because the chance of finding a sample with a very high intensity is low, it is unlikely to observe samples beyond some amplitude within the finite observing time. Finally, under the assumption of a uniform spatial distribution of RFI transmitters, samples with very high amplitude would have to be produced by transmitters that are very close to the telescope. However, it is likely that the uniform spatial distribution of transmitters will break down at closer distances.

@@

With the assumption that $Ig/r_L^2\sim$ a uniform distribution, the area labelled in Fig.~\ref{fig:explanation-constraints} as $N_b$ is also part of the RFI distribution, and a stronger constraint $\tilde S_L$ can be found by solving
\begin{equation}
 \int\limits_{\tilde S_L}^{S_U} g(S) dS + \tilde S_{L} g(\tilde S_{L}) = N_\textrm{total} - N_\textrm{part},
\end{equation}
which yields
\begin{equation} \label{eq:lower-limit-2}
\tilde S_L = \sqrt[\alpha+1]{ - \frac{1}{\alpha} \left( \frac{\alpha+1}{\beta} \left(N_\textrm{part} - N_\textrm{total}\right) + S_1^{\alpha+1} \right) }.
\end{equation}

@@

\section{Data description} \label{sec:dist-data}
We have analysed the distributions of two data sets. Both data sets are 24 h LOFAR RFI surveys and are extensively analysed in
\citet{lofar-radio-environment}. % TODO
In one set, the low-band antennas (LBA) were used and the frequency range 30.1--77.5~MHz was recorded, while in the other the high-band antennas (HBA) were used to record the frequency range 115.0--163.3~MHz. More stations were used in the LBA set. The specifications of the two sets are listed in Table~\ref{table:dist-data-specs}. The stations that have been used are geometrically spread over an area of about 80~km and 30~km in diameter at maximum for the LBA and HBA sets respectively.

@@

\begin{table}
\caption{Data set specifications}\label{table:dist-data-specs}
\begin{center}
\begin{tabular}{lrr}
                    & \textbf{LBA set}& \textbf{HBA set} \\
\hline
\hline
Observation date    & 2011-10-09      & 2010-12-27 \\
Start time          & 06:50 UTC       & 0:00 UTC \\
Length              & 24 h           & 24 h \\
Time resolution     & 1 s             & 1 s \\
\hline
Frequency range     &  30.1--77.5 MHz & 115.0--163.3 MHz\\
Frequency resolution & 0.76 kHz    & 0.76 kHz \\
Number of stations  &  33           & 13 \\
Total size          & 96.3 TiB        & 18.6 TiB \\
\hline
Field               & NCP             & NCP \\
Amount RFI detected & 1.77\%       & 3.18\% \\
\hline
\hline
\end{tabular}
\end{center}
\end{table}

@@

If we go back to Fig.~\ref{fig:histogram-passband-corrected}, we see that in the LBA the power law is stable for about three orders of magnitude, and one order more in the HBA. Fig.~\ref{fig:plot-slopes} shows the slope of the log-log plot as a function of amplitude, which was constructed by performing linear regression in a sliding window, with a window size of 1 decade. The HBA shows very little structure in the slope, but the LBA is less stable and shows some features in its power-law part. Linear regression on the power-law part of the log-log plot results in a slope of $-1.62$ for the LBA and $-1.53$ for the HBA. These and the other derived quantities have been summarized in Table~\ref{table:dist-data-quantities}. Although the HBA slope does not show any other significant features besides the Rayleigh and power-law curves, the LBA power law ends with a bulge around an amplitude of $10^6$. This bulge is caused by a very strong source affecting lots of samples, and is a single outlier in the spatial distribution. We found this is caused by RFI observed for about an hour in the late afternoon in the lower LBA frequency regime, around $30$--$40$~MHz. Leaving this frequency range out flattens the bulge significantly, but does not completely eliminate it, because the source turned the receivers to a non-linear state, causing leakage at lower intensity levels in the other sub-bands. Unlike linear regression, the fitting region of the Hill estimator is not limited at the high end. Consequently, because of the bulge, the Hill estimator evaluates for the LBA into a slope that is less steep, with a value of $-1.53$. For the HBA set, the Hill estimator is equal to the $-1.53$ value found by linear regression.

@@

Because some RFI sources might be stationary, the signals from these sources will add coherently over time. Therefore, the amount that time integration can lower the flux density of RFI might be limited. Additionally, some RFI sources are received by multiple stations of the array, and by multiple feeds of the individual antennas. Therefore, integrating data from different interferometers and data from the two polarized feeds might not lower the noise level that is caused by RFI at some point. In summary, RFI is unlike normal noise and might be coherent over time, interferometer and feed.

@@

\subsection{Interference-reducing effects} \label{sec:coherence-reduction}
When integrating data, it is likely that the actual noise limit will be significantly lower than the given upper limit, which was determined on high resolution. There are several reasons for this, which we will summarize one by one.
\begin{itemize}
 \item Many RFI sources have a variable geometric phase, because they move or because their path of propagation changes. This would cause them to sum (partly) incoherently over time, and thus go approximately down with the noise.
 \item Many RFI sources will be seen by only a few stations. This would make the histogram go down more quickly at lower amplitudes.
 \item For the shortest baselines at 150~MHz, the far field starts around 1~km. Some RFI sources will therefore be in the near field, especially for the longer baselines. In that case, a source will not add up coherently over the interferometers that see the particular source, as the interferometers see them with different phases.
 \item Many stationary sources are not constant over time, thus will be attenuated somewhat by averaging.
 \item We have assumed 100\% of the spectrum is occupied by RFI. If, say, only 10\% of the spectrum is occupied, the expected value of the leaked RFI level decreases with about $25$\%, and if the detected 2.68\% true-positives contain all RFI, there is no leaked RFI at all. With current data, one can only speculate how much the electromagnetic spectrum is truly occupied.
 \item Stationary RFI sources in a uniform spatial distribution will interfere both constructively and destructively with each other. Individually, the sources will have fixed geometric phases, but because the baselines are much longer than the wavelength, their phases will become uniformly spread. Therefore, they will add incoherently.
\end{itemize}

@@

With the current strategy, it is likely that the LOFAR EoR project will encounter some RFI on some frequencies when averaging lots of observing nights, although this still has to be seen. To mitigate this leaked RFI, the detection can be executed at higher signal-to-noise levels. The current results indicate that a lot of RFI is not coherent, and the situation is promising. Considering the current RFI results, and the availability of further mitigation steps, it is very likely to assume that RFI will not be problematic for the detection Epoch of Reionisation with LOFAR.

>>>>

\begin{abstract}
Feeble undetected sources of radio-frequency interference (RFI) might become visible after long time integration of radio observations if they are coherently present over time. Thereby, they might obstruct the detection of the weak astronomical signals of interest. This issue is especially important for Epoch of Reionisation (EoR) projects that try to detect the faint redshifted HI signals from the time of the earliest objects in the Universe. We explore the low-frequency RFI situation by studying the $\log N$--$\log S$ histograms of data observed with LOFAR. The histograms shows an RFI distribution that follows a power-law distribution with an estimated exponent around -1.5. With several assumptions, this can be explained with a uniform distribution of terrestrial radio sources whose radiation follows existing propagation models. Extrapolation of the power law would imply that the current LOFAR EoR observations should be severely RFI limited if these sources are coherently present. Since these observations have achieved the thermal (sky) noise limit, we conclude that it is probable that most of the RFI sources do not add coherently.
\end{abstract}

@@

\section{Introduction}
Radio astronomy concerns itself with the observation of radiation from celestial sources at radio wavelengths. However, astronomical radio observations can be affected by radio-frequency interference (RFI), which makes it difficult to calibrate the instrument and achieve high sensitivities \citep{impact-of-warc79,interference-and-radioastronomy-1991,interference-model-lemmon,rfi-mitigation-overview-fridman-baan}. Many techniques have been designed to mitigate this effect, such as detection and flagging of data \citep{chi-square-time-blanking-weber, multichannel-rfi-mitigation, exoplanet-detection-with-rfi, wsrt-rfims, pulse-blanking, effelsberg-rfi-mitigation, LOFAR-RFI-pipeline}, adaptive cancellation techniques \citep{adaptive-cancellation,post-correlation-reference-signal} and spatial filtering \citep{multichannel-rfi-mitigation, ellingson-spatial-nulling-2002, hampson-spatial-nulling-2002, boonstra-dissertation, spatial-filtering-parkes-multibeam-for-pulses, post-correlation-filtering}. In most cases, these methods excise enough of the interference to allow calibration and imaging of the data.

Typical radio observations record a few hours of data, and the results are integrated. In these cases, it often suffices to only excise interference that is apparent and thus above the noise, to reach the thermal noise limit of the instrument. A new challenge arises, however, when one desires much deeper observations, and hundreds of hours of observations need to be integrated. In such a case, feeble interference caused by stationary sources might not manifest itself above the noise in individual observations, but might be coherently present in the data. Subsequently, when averaging these data, the interference might become apparent and occlude the signal of interest. This is very relevant for the 21-cm Epoch of Reionisation (EoR) experiments, because they involve long integration times. Several such experiments are underway, to either measure the angular power spectrum \citep{gmrt-eor-2011-paciga,de-bruyn-eor-ursi-2011,eor-paper-2011-jacobs,eor-mwa-2012-williams} or the global signal \citep{edges}. For these experiments, it is important to know the possible effect of low-level interferences on the data, as these might overshadow or alter the signal of interest. In this article, we will connect new insights about RFI to the angular Epoch of Reionisation experiment that is using the Low-Frequency Array (LOFAR) \citep{de-bruyn-eor-ursi-2011}.

This work explores the information that is present in interference distributions, in order to analyse possible low-level interference that is not detectable by standard detection methods. Our approach is similar to the radio source count analysis that is used in cosmology \citep{vonhoerner-radio-source-counts}, also named $\log N$ -- $\log S$ analysis, where $N$ and $S$ refer to the celestial source count and brightness respectively. The slope in such a plot contains information about source populations, their luminosity functions and the geometry of the Universe. We analyse such a double-logarithmic plot for the case of terrestrial sources, with the ultimate goal of predicting their full spatial and brightness distributions. This results in a better insight in the effects of low-level interference and allows one to simulate the effects of interference more accurately. A distribution can be expressed either as a differential or cumulative function. This work uses differential distributions, because they are better suited for visualizing the parameters we are interested in, although the two distributions basically convey the same information.

@@

Next, we assume that all RFI sources have equal constant strength $I$ and follow a uniform spatial distribution in a two-dimensional plane. This is obviously a simplification, as the RFI sources are actually distributed on the surface of a sphere with different heights. Beyond some distance, the Earth will partly block the path between transmitter and receiver. Therefore, the assumption of uniformly distributed sources on a two-dimensional plane is only approximately valid. Using these assumptions, we can express the expected cumulative frequency density of sources $F_{\textrm{distance} \le r}(r)$ at distance $r$ as
\begin{equation} \label{eq:cumulative-distance-function}
F_{\textrm{distance} \le r}(r)=c\pi r^2,
\end{equation}
for some constant $c$ that represents the number of sources per unit area. In other words, we will observe $F$ sources that are at most a distance $r$ away.

@@

Converting from a subtracted term in decibels to a flux density factor $L_S$, and collecting the terms of Eq.~\eqref{eq:hata-model} that are not depending on $r$ in a single variable $\zeta$, results in
\begin{equation} \label{eq:propagation-loss-factor}
 L_S = \zeta r^{4.49 - 0.655 \log_{10} h_b},
\end{equation}
with
\begin{equation}
 \zeta = \frac{f_c^{2.616}}{h_b^{1.382}} - 10^{6.955-\frac{1}{10} a(h_m)}.
\end{equation}
Note that according to Hata's model, the exponent of the power law in Eq.~\eqref{eq:propagation-loss-factor} depends only on the height of the transmitting antenna, i.e., it is independent of frequency, receiver height and urban density. Now, if in Eqs.~\eqref{eq:cumulative-rfi-distribution} and \eqref{eq:two-dimensional-distribution} one replaces the definition of $S(r)$ from Eq.~\eqref{eq:amplitude-fall-off-in-free-space} with one that includes the propagation effects,
\begin{equation}
 S(r) = \frac{\zeta Ig}{r^\eta},
\end{equation}
with $\eta = 0.655 \log_{10} h_b - 4.49$, one finds the frequency density function $f_p$ that considers propagation effects,
% S(r):=I*g/(r^alpha);
% assume(g>0,I>0,S>0);
% diff(c*%pi*r^2,S),solve(S(r)=S,r);
\begin{align}
f_p(S)
= & \frac{d}{dS} c \pi \left( \frac{\zeta Ig }{S} \right)^{2/\eta}
= \frac{c 2\pi}{\eta S} \left( \frac{\zeta Ig}{S} \right)^{2/\eta}.
\end{align}
Consequently, due to non-free space propagation effects, the observed log-log histogram is predicted to have a $\frac{2}{\eta}-1$ slope. By substituting $\eta$, one finds
\begin{equation}
\textrm{slope}(h_b) = \frac{1}{0.3275 \log_{10} h_b - 2.245} - 1.
\end{equation}
\begin{figure}
\begin{center}\hspace{-1cm}\includegraphics[width=6cm]{img/transmitter-height/plot-transmitter-height-trimmed}
\caption{Effect of transmitter height on the slope of a log-log histogram.}
\label{fig:plot-transmitter-height}
\end{center}
\end{figure}
This is valid for transmitters that have a height of 30--200 m, the range over which Hata's model was defined. This yields estimated distribution slopes of $-1.57$ and $-1.67$ for $30$ m and $200$ m high transmitters respectively. In Figure~\ref{fig:plot-transmitter-height}, the slope value is plotted as a function of the transmitter height, including extrapolated values for transmitter heights down to 1 m.

@@

To calculate the corresponding amplitude distribution that is formed by the combination of the real and imaginary component distributions, it is easier to perform numerical simulations, as analytic derivation is not trivial. The histogram can be numerically estimated by drawing complex real and imaginary samples from the two distributions and calculating and counting the amplitudes. A sample can be drawn from the RFI distribution by normalizing and inverting the cumulative frequency function in Eq.~\eqref{eq:cumulative-rfi-distribution} and evaluating it for a uniformly distributed variable. Note however that $F_{\textrm{amplitude}\le S}(S)$ is not limited; when decreasing the amplitude $S$ towards zero, the number of sources will go to infinity. Therefore, to draw a sample by using the inversion method, this would require a uniformly distributed variable $\sim U(0,\infty)$. By assuming that there are no samples beyond some limiting distance $r_L$, then $F_{\textrm{amplitude}\le S}(S) = N-c \pi r_L^2$ for $S < Ig/r_L^2$. Consequently, by sampling $x_u$ from a uniform distribution with $0 \le x_u \le N-c \pi r_L^2$ and applying the inverse of Eq.~\eqref{eq:cumulative-rfi-distribution}, we will sample amplitudes $S$ with $S \ge Ig/r_L^2$, and $S\sim$ a power-law distribution.

@@

\subsection{Determining RFI distribution limits} \label{sec:rfi-distribution-constraints}
In this section we will show methods to find $S_U$ and $S_L$, the upper and lower flux limits of the power-law distribution at which the power law breaks down. Once the exponent of the power-law part of the distribution is estimated with the previously discussed techniques, the distribution can be extrapolated to find a lower flux limit. Assume that we have found a power law with exponent $\alpha$ and factor $\beta$ over an amplitude region $[S_1 ; S_2]$, resulting in the following frequency density function $h$:
\begin{equation} \label{eq:amplitude-power-law}
h(S) = \beta S^\alpha.
\end{equation}
\begin{figure}
\begin{center}
\includegraphics[width=7.5cm]{img/explanation-lower-constraint/FindingCounts}
\includegraphics[width=7.5cm]{img/explanation-lower-constraint/FindingConstraints}
\caption{Cartoon of how a constraint on the lower fall-off point of the power-law distribution can be determined. Note that the labelled areas are areas as occupied in a linear plot, i.e., the integration of the density function. Areas in a log-log plot are not linearly related to the integral. There are two ways to estimate the lower constraint $S_L$: (i) the areas $N_a$ and $N_\textrm{total}-N_\textrm{part}$ are equal if $Ig/r^2$ is constant during the observation, and (ii) if one assumes $Ig/r^2\sim$ uniform, then $N_a+N_b = N_\textrm{total}-N_\textrm{part}$.}
\label{fig:explanation-constraints}
\end{center}
\end{figure}

Assume also that the histogram contains $N_\textrm{part}$ (RFI) samples with amplitude $> S_1$, as sketched in Fig.~\ref{fig:explanation-constraints}. The hypothetical upper limit $S_U$ of the distribution can now be found, i.e., the highest amplitude that would be observed when the distribution follows the power law up to that point, by solving
\begin{equation} \label{eq:upper-limit-constrained}
 \int\limits_{S_1}^{S_U} h(S) dS = N_\textrm{part}.
\end{equation}
In practice, the observed histogram will break down beyond some amplitude because of several reasons. First and most importantly, the samples themselves are digitized with an analogue-to-digital converter (ADC) with limited range. Second, we observe for a limited time and the frequency count is discrete. Because the chance of finding a sample with a very high intensity is low, it is unlikely to observe samples beyond some amplitude within the finite observing time. Finally, under the assumption of a uniform spatial distribution of RFI transmitters, samples with very high amplitude would have to be produced by transmitters that are very close to the telescope. However, it is likely that the uniform spatial distribution of transmitters will break down at closer distances.

@@

With the assumption that $Ig/r_L^2\sim$ a uniform distribution, the area labelled in Fig.~\ref{fig:explanation-constraints} as $N_b$ is also part of the RFI distribution, and a stronger constraint $\tilde S_L$ can be found by solving
\begin{equation}
 \int\limits_{\tilde S_L}^{S_U} h(S) dS + \tilde S_{L} h(\tilde S_{L}) = N_\textrm{total} - N_\textrm{part},
\end{equation}
which yields
\begin{equation} \label{eq:lower-limit-2}
\tilde S_L = \sqrt[\alpha+1]{ - \frac{1}{\alpha} \left( \frac{\alpha+1}{\beta} \left(N_\textrm{part} - N_\textrm{total}\right) + S_1^{\alpha+1} \right) }.
\end{equation}

@@

\section{Data description} \label{sec:dist-data}
We have analysed the distributions of two data sets. Both data sets are 24 h LOFAR RFI surveys and are extensively analysed in
\citet{lofar-radio-environment}. % TODO
In one set, the low-band antennas (LBA) were used and the frequency range 30.1--77.5~MHz was recorded, while in the other the high-band antennas (HBA) were used to record the frequency range 115.0--163.3~MHz. More stations were used in the LBA set. The specifications of the two sets are listed in Table~\ref{table:dist-data-specs}. The stations that have been used are geometrically spread over an area of about 80~km and 30~km in diameter at maximum for the LBA and HBA sets respectively. For EoR detection experiments, the HBA are more important than the LBA, because they cover the frequency range of the redshifted EoR signal.

@@

\begin{table}
\caption{Data set specifications}\label{table:dist-data-specs}
\begin{center}
\begin{tabular}{lrr}
                    & \textbf{LBA set}& \textbf{HBA set} \\
\hline
\hline
Observation date    & 2011-10-09      & 2010-12-27 \\
Start time          & 06:50 UTC       & 0:00 UTC \\
Length              & 24 h           & 24 h \\
Time resolution     & 1 s             & 1 s \\
\hline
Frequency range     &  30.1--77.5 MHz & 115.0--163.3 MHz\\
Frequency resolution & 0.76 kHz    & 0.76 kHz \\
Number of stations  &  33           & 13 \\
Total size          & 96.3 TB        & 18.6 TB \\
\hline
Field               & NCP             & NCP \\
Amount RFI detected & 1.77\%       & 3.18\% \\
\hline
\hline
\end{tabular}
\end{center}
\end{table}

@@

If we go back to Fig.~\ref{fig:histogram-passband-corrected}, we see that in the LBA the power law is stable for about three orders of magnitude, and one order more in the HBA. Fig.~\ref{fig:plot-slopes} shows the slope of the log-log plot as a function of amplitude, which was constructed by performing linear regression in a sliding window, with a window size of 1 decade. The HBA shows very little structure in the slope, but the LBA is less stable and shows some features in its power-law part. Linear regression on the power-law part of the log-log plot results in a slope of $-1.62$ for the LBA and $-1.53$ for the HBA. These and the other derived quantities have been summarized in Table~\ref{table:dist-data-quantities}. Although the HBA slope does not show any other significant features besides the Rayleigh and power-law curves, the LBA power law ends with a bulge around an amplitude of $10^6$. This bulge is caused by a very strong source affecting lots of samples, and is a single outlier in the spatial distribution. We found this is caused by RFI observed for about an hour in the late afternoon in the lower LBA frequency regime, around $30$--$40$~MHz. Leaving this frequency range out flattens the bulge significantly, but does not completely eliminate it, because the source put the receivers in a non-linear state, causing leakage at lower intensity levels in the other sub-bands. Unlike linear regression, the fitting region of the Hill estimator is not limited at the high end. Consequently, because of the bulge, the Hill estimator evaluates for the LBA into a slope that is less steep, with a value of $-1.53$. For the HBA set, the Hill estimator is equal to the $-1.53$ value found by linear regression.

@@

Because some RFI sources might be stationary, the signals from these sources will add coherently over time. Therefore, the amount that time integration can decrease the flux density of RFI might be limited. Additionally, some RFI sources are received by multiple stations of the array, and by multiple feeds of the individual antennas. Therefore, integrating data from different interferometers and data from the two polarized feeds might not lower the noise level that is caused by RFI at some point. In summary, RFI is unlike normal noise and might be coherent over time, interferometer and feed.

@@

\subsection{Interference-reducing effects} \label{sec:coherence-reduction}
When integrating data, it is likely that the actual noise limit from low-level RFI will be significantly lower than the given upper limit, which was determined at highest LOFAR resolution, before further averaging. There are several reasons for this, which we will summarize one by one.
\begin{itemize}
 \item Many RFI sources have a variable geometric phase, because they move or because their path of propagation changes. This would cause them to sum (partly) incoherently over time, and thus go approximately down with the noise.
 \item Many RFI sources will be seen by only a few stations. This would make the histogram go down more quickly at lower amplitudes.
 \item For the shortest baselines at 150~MHz, the far field starts around 1~km. Some RFI sources will therefore be in the near field, especially for the longer baselines. In that case, a source will not add up coherently over the interferometers that see the particular source, as the interferometers see them with different phases.
 \item Many stationary sources are not constant over time, thus will be attenuated somewhat by averaging.
 \item We have assumed 100\% of the spectrum is occupied by RFI. If, say, only 10\% of the spectrum is occupied, the expected value of the leaked RFI level decreases with about $25$\%, and if the detected 2.68\% true-positives contain all RFI, there is no leaked RFI at all. With current data, one can only speculate how much the electromagnetic spectrum is truly occupied.
 \item Stationary RFI sources in a uniform spatial distribution will interfere both constructively and destructively with each other. Individually, the sources will have fixed geometric phases, but because the baselines are much longer than the wavelength, their phases will become uniformly spread. Therefore, they will add incoherently.
\end{itemize}

@@

With the current strategy, it is likely that the LOFAR EoR project will encounter some RFI on some frequencies when averaging lots of observing nights, although this still remains to be seen. To mitigate this leaked RFI, the detection can be executed at higher signal-to-noise levels. The current results indicate that a lot of RFI is not coherent, and the situation is promising. Considering the current RFI results, and the availability of further mitigation steps, we conclude that RFI will likely not be problematic for the detection Epoch of Reionisation with LOFAR.
